---
title: "Scraping CRAN"
author: "Julia Silge"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(cache = TRUE, echo = TRUE,
                      warning = FALSE, message = FALSE, 
                      dpi = 180, fig.path = "figs/")
options(width=80)
library(ggplot2)
library(silgelib)
theme_set(theme_roboto())
library(tidyverse)
```

This script takes a while to run because it queries a webpage for every package in the CRAN archives.

## Get the data

Download all currently available packages on CRAN.

```{r pkgs}
library(rvest)
library(stringr)
library(lubridate)

## which CRAN mirror am I going to use...
CRAN_page <- function(...) {
    file.path("https://cran.rstudio.com/src/contrib", ...)
}

pkgs_raw <- read_html(CRAN_page()) %>% 
    html_nodes("table") %>% 
    .[[1]] %>%
    html_table()

pkgs_raw <- pkgs_raw[,-1]

pkgs <- pkgs_raw %>%
    filter(Size != "-",
           str_detect(Name, "tar.gz$")) %>%
    mutate(Date = dmy_hm(`Last modified`),
           Name = str_extract(Name, "^[^_]+(?=_)")) %>%
    select(-Size, -Description) %>%
    as_tibble()

pkgs
```

Download the archived packages on CRAN, getting their first available dates and how many versions.

```{r archives, dependson="pkgs"}
archives_raw <- read_html(CRAN_page("Archive")) %>% 
    html_nodes("table") %>% 
    .[[1]] %>%
    html_table()

archives_raw <- archives_raw[,-1]

archives_processed <- archives_raw %>%
    filter(str_detect(Name, "/$")) %>%
    mutate(Date = dmy_hm(`Last modified`),
           Name = str_sub(Name, end = -2)) %>%
    select(-Size, -Description) %>%
    as_tibble()


read_page <- function(name) {
    message(name)
    read_html(CRAN_page("Archive", name)) %>% 
        html_nodes("td") %>% 
        html_text()
}

archives_scraped <- archives_processed %>%
    mutate(page = map(Name, read_page))

archives <- archives_scraped %>%
    mutate(Date = dmy_hm(map_chr(page, ~ .[8])),
           ArchivedVersions = map_dbl(page, ~ length(.) / 5 - 1)) %>%
    select(-page)

archives
```

## Join the data

Now put it all together.

- Packages that are in `archives` but not `pkgs` are no longer on CRAN.
- Packages that are in `pkgs` but not `archives` only have one CRAN release.
- Packages that are in both dataframes have had more than one CRAN release.

```{r all_pkgs, dependson=c("archives", "pkgs")}
all_pkgs <- bind_rows(archives %>% 
                          anti_join(pkgs, by = "Name") %>% 
                          mutate(Archived = TRUE),
                      pkgs %>% 
                          anti_join(archives, by = "Name") %>% 
                          mutate(ArchivedVersions = 0,
                                 Archived = FALSE),
                      archives %>%
                          semi_join(pkgs, by = "Name") %>%
                          mutate(Archived = FALSE)) %>%
    mutate(Versions = ifelse(Archived, ArchivedVersions, ArchivedVersions + 1)) %>%
    arrange(Name)

all_pkgs
```

Now let's add up how many have been available, by month.

```{r counts, dependson="all_pkgs", fig.width=8, fig.height=6}
all_pkgs %>%
    filter(!Archived) %>%
    group_by(Date = floor_date(Date, unit = "month")) %>%
    summarise(NewPackages = n()) %>%
    ungroup %>%
    mutate(TotalPackages = cumsum(NewPackages)) %>%
    ggplot(aes(Date, TotalPackages)) +
    geom_line(size = 1.5, alpha = 0.8, color = "midnightblue") +
    labs(x = NULL, y = "Number of available packages",
         title = "How many packages are available on CRAN?",
         subtitle = "Only packages that are still available")
```


```{r archived, dependson="all_pkgs", fig.width=6, fig.height=6}
all_pkgs %>%
    ggplot(aes(Archived)) +
    geom_histogram(stat = "count", alpha = 0.8, fill = "midnightblue") +
    scale_x_discrete(labels=c("Still available", "Archived, no longer available")) +
    labs(y = "Number of packages", x = NULL,
         title = "How many packages are no longer available on CRAN?",
         subtitle = "About 10% of total packages are no longer available")
         
```


```{r versions, dependson="all_pkgs", fig.width=8, fig.height=6}
all_pkgs %>%
    ggplot(aes(Versions)) +
    geom_histogram(binwidth = 10, alpha = 0.8, fill = "midnightblue") +
    labs(y = "Number of packages", x = "Number of versions on CRAN",
         title = "How many versions do CRAN packages have?",
         subtitle = "About 25% of packages are on their first version")
    
```

